Data Specialization for a Staged Lambda Calculus

### 2014-11-09 paper/semantics.tex
complete rewrite of section on the language and semantics

@@

The relation 1-evaluation, defined by $\redonesym$ judgment as shown
in \ref{fig:diaSemantics}, takes an stage \bbone\ term%
\footnote{ The input to $\redonesym$ may be open on stage \bbtwo\ variables, but
the type system ensures that those must occur under a $\next$. Consequently
$\redonesym$ will never directly encounter a variable.}
%
to a residual table $\xi$ and a partial value $v$.  For non-staging
features of \lang, $\redonesym$ is essentially standard call-by-value
evaluation but in addition gathers the subterms' residual tables into
a single table. \ur{The syntax of residual tables should be defined in
  the terminology figure} Residual table entries are created by the
evaluation of \next: 1-evaluation reduces a \next expression by
reducing the body of the \next with 2-evaluation to obtain a
residual $q$ and returns a residual table that maps a freshly
generated variable to the residual.  \ur{Variable should be fresh in
  the rule.}

The relation 2-evaluation, defined by $\redtwosym$ judgment, as shows
in \figref{diaSemantics}, takes a stage \bbtwo\ term and reduces it to
a residual.  Apart from the evaluation of \prev and \pause, the
evaluation is straightforward. To reduce a \prev term, 2-evaluation
reduces the body by 1-evaluation to obtain a residual table and a
partial value.  It then reifies the residual table and the partial
value into an residual term, by creating for each variable-residual
pair in the table a corresponding variable binding in the residual,
finally ending with the partial value.
\ur{There seems to be errors in the semantics in the handling of
  recursive types.  I don't see any reason to do any reductions here.
If you want to, the description above needs to be updated.}

...UMUT: UP TO HERE...

 In addition, 1-evaluation {\em reifies} the contents of
the residual table into a residual when changing from stage-2 to stage

 The residual
table implements the explicit substitutions in our example---it maps fresh
stage \bbtwo\ variables (which may appear inside $\next$ blocks in $v$) to
residuals.

In order to avoid duplicating subterms, evaluation collects explicit
substitutions into a single {\em residual table} that is also the output of
evaluation.

 the auxiliary $\reifysym$ (reification). 
Along with these are three judgments identifying forms of values:
$\pvalsym$ (partial values), $\ressym$ (residuals), 
and the auxiliary $\fcon$ (residual table).
These judgments are related as follows:
\begin{itemize}
\item Evaluation sends a stage~\bbone\ term to a partial value.

\item Specialization sends a stage~\bbtwo\ term to a residual.

\item Reification sends a single residual and a residual table to another residual.

\item Evaluation depends on specialization for $\next$ terms, which have
stage~\bbtwo\ subterms.
\item Specialization depends on evaluation for $\prev$ and $\pause$, which have
stage~\bbone\ subterms. The residual table produced by evaluation is also
reified here.
\end{itemize}

The relation 2-evaluation takes a stage-2 term and maps to a residual.

To generate the residual table, the dynamics semantics relies on a
{\em reification} relation, $\reifysym$, that maps a table to a
residual, essentially generating a target expression representing the
computations in the table.

We can evaluate a \lang\ term as follows: we first apply 1-evaluation
to obtain a partial value and a residual table, we then construct a
2-term from the partial value and the residual table by using
reification, and then reduce this term to a residual via 2-evaluation.

Thus, the dynamic semantics for \lang\ evaluates all of a term's stage \bbone\
subexpressions before any of its stage~\bbtwo\ subexpressions. This results in
a stage \bbtwo\ term with no stage~\bbone\ content, 
which can be described as a term in a monostaged language called \langTwo. 
Then, at stage~\bbtwo, we complete
evaluating this term with a standard dynamic semantics, $\tworedsym$. (The
rules for this judgment are not shown, but they are standard.)

\crem{Can we cut \ref{fig:diaSemanticsSpec} from the paper? Perhaps just put a
few representative rules into \ref{fig:diaSemantics}?}

At $\next~e$, in contrast, $\redonesym$ specializes into $e$ for stage~\bbone\ 
subexpressions to evaluate.  This is implemented by the $\redtwosym$ judgment,
which takes a stage \bbtwo\ term to a residual. 
For the most part (\ref{fig:diaSemanticsSpec}), specialization simply recurs into
all subexpressions; at $\prev$, however, it resumes $\redonesym$ evaluation,
which produces a residual table $\xi$ and (by canonical forms) an expression
$\next\{\hat y\}$. $\xi$ is then reified using $\reifysym$ into a series of let
bindings enclosing $\mathtt{\hat y}$.
Once specialization reduces $e$ to a residual $q$, the output of $\redonesym$ is a
fresh variable wrapped in a $\next$ block ($\next~\hat y$), along with a residual
table which maps that variable to the residual ($\hat y \mapsto q$).

Within specialization lies a subtle---if perhaps unintuitive---feature.  
Observe that specialization will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
statement in its search for stage~\bbone\ code. 
Thus the evaluation of that stage~\bbone\ code will occur {\em regardless of the eventual value of the predicate},
and so a term like 
\begin{lstlisting} 
1`next{
  `2`if true 
  then hold{`1`3+4`2`} 
  else prev{`1`spin() (* loops forever *)`2`}`1`
}`
\end{lstlisting}
will fail to evaluate at stage \bbone.
Although it may seem undesirable, this behavior is a deliberate feature.
Indeed it is critical to the quickselect example, 
wherein all three branches of the $n$-$k$ comparison are evaluated at stage~\bbone.
As always, it is the programmer's responsibility to ensure that her staging annotations 
do not cause a program to loop forever.

The context ($\Gamma$) keeps track of stage \bbtwo\ variables in the input term. 
These both appear in the original program at stage \bbtwo\ and are inserted by the semantics.

As an optimization, we can include the special-case rule,
\begin{mathpar}
\inferdiaone [hat] {\red {\next~\hat y}{\cdot,\next~\hat y}}{\cdot}
\end{mathpar}
to avoid one-for-one variable bindings in the residual.

\subsubsection {Top-Level Evaluation}
\label{sec:partialeval}

% KAYVONF: good statement, but hold out for now
%The hope of partial evaluation is that $f_x$ is cheaper to execute than $f$, meaning that we can save work if we must %evaluate it many times.

These dynamics allow us to define a partial evaluator for \lang\ by identifying
static with stage \bbone\ and dynamic with stage \bbtwo.  Specifically, we
encode $f$ as a \lang\ expression with a function type of the form
$A\to\fut(B\to C)$.%
%\cprotect\footnote{We can rewrite \texttt{fexp} in this form, or simply apply
%the following higher-order function which makes the adjustment:
%\begin{lstlisting} 
%let adjust (f : $int * int -> $int) =
%  fn (p : int) => 
%    next{
%      fn (b : int) => 
%        prev{f (next {b}, p)}
%    }
%\end{lstlisting}}
%
Here $A$ is the static input, $B$ is the dynamic input, and $C$ is the output.

Once a stage \bbone\ argument $a:A$ is provided, we can evaluate the partially-applied
function:
$\cdot\vdash f~a \mathop{\redonesym} [\xi,v]$.
The result is an environment $\xi$ and a partial value $v$ of type $\fut(B\to
C)$, which by canonical forms must have the form $v = \next~\hat y$. 
Next, we reify this environment into a sequence of \verb|let|-bindings
enclosing $\hat y$, via $\reify\xi{\hat y}{f_a}$. 
Because reification preserves types, the resulting residual $f_a$ has type $B\to C$ in \langTwo, so we can apply it to some $b:B$
and compute the final result of the function, $f_a~b \mathop{\tworedsym} c$.

That this sequence of evaluations is in fact staged follows from our
characterizations of partial values and residuals, that $\redonesym$
outputs a partial value, and that $\reifysym$ outputs an expression in \langTwo.

For example, to specialize quickselect to a particular list, 
\begin{lstlisting}
qsStaged [5,2,7,4,1]
\end{lstlisting}
\TODO finish example

>>>>

\input{figures/dynamics}

@@

The relation 1-evaluation, defined by $\redonesym$ judgment as shown
in \ref{fig:diaSemantics}, takes an stage \bbone\ term%
\footnote{ The input to $\redonesym$ may be open on stage \bbtwo\ variables, but
the type system ensures that those must occur under a $\next$. Consequently
$\redonesym$ will never directly encounter a variable.}
%
to a residual table $\xi$ and a partial value $v$.  For non-staging
features of \lang, $\redonesym$ is essentially standard call-by-value
evaluation but in addition gathers the subterms' residual tables into
a single table. \ur{The syntax of residual tables should be defined in
  the terminology figure} Residual table entries are created by the
evaluation of \next: 1-evaluation reduces a \next expression by
reducing the body of the \next with 2-evaluation to obtain a
residual $q$ and returns a residual table that maps a freshly
generated variable to the residual.

The relation 2-evaluation, defined by $\redtwosym$ judgment, as shows
in \figref{diaSemantics}, takes a stage \bbtwo\ term and reduces it to
a residual.  Apart from the evaluation of \prev\ and \pause, the
evaluation is straightforward. To reduce a \prev term, 2-evaluation
reduces the body by 1-evaluation to obtain a residual table and a
partial value.  It then reifies the residual table and the partial
value into an residual term, by creating for each variable-residual
pair in the table a corresponding variable binding in the residual,
finally ending with the partial value.

We can use the evaluation relations in the dynamic semantics to
evaluate a term to evaluate a \lang\ term as follows. We first apply
1-evaluation to obtain a partial value and a residual table. We then
construct a 2-term from the partial value and the residual table by
using reification, and then reduce this term to a residual via
2-evaluation.  Finally, we evaluate the residual, which is a term in a
our monostaged target language \langmono, to a value by using  a
standard evaluation relation for \langmono.
